---
title: "Profile likelihood confidence intervals"
author: '[Francisco Javier Rubio](https://sites.google.com/site/fjavierrubio67/)'
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Profile likelihood confidence intervals for the parameters of the normal distribution

Let ${\bf x} = (x_1,\dots,x_n)$ be a sample from a distribution with density $f(\cdot;\boldsymbol{\theta})$. Suppose that $\boldsymbol{\theta} = (\boldsymbol{\delta},\boldsymbol{\xi})$, where $\boldsymbol{\delta}$ is a parameter of interest, and $\boldsymbol{\xi}$ is a nuisance parameters. Let:
$$
L(\boldsymbol{\theta}\mid {\bf x}) = L(\boldsymbol{\delta},\boldsymbol{\xi} \mid {\bf x}) = \prod_{i=1}^{n} f(x_i;\boldsymbol{\theta}),
$$
be the likelihood function of $\boldsymbol{\theta}$. Let $\widehat{\boldsymbol{\theta}} = (\widehat{\boldsymbol{\delta}},\widehat{\boldsymbol{\xi}})$ be the MLE of $\boldsymbol{\theta}$. Then, the profile likelihood (or profile likelihood ratio) of $\boldsymbol{\delta}$ is:
$$
R(\boldsymbol{\delta}\mid {\bf x}) = \dfrac{L(\boldsymbol{\delta},\widehat{\boldsymbol{\xi}}(\boldsymbol{\delta}) \mid {\bf x})}{L(\widehat{\boldsymbol{\theta}}\mid {\bf x})} = \dfrac{\max_{\boldsymbol{\xi}} L(\boldsymbol{\delta},\boldsymbol{\xi} \mid {\bf x}) }{\max_{\boldsymbol{\theta}}L({\boldsymbol{\theta}}\mid {\bf x})},
$$
where $\max_{\xi} L(\boldsymbol{\delta},\boldsymbol{\xi} \mid {\bf x})$ means that we are maximising the likelihood with respect to $\boldsymbol{\xi}$, for a fixed value of $\boldsymbol{\delta}$. Then, profile likelihood is bounded $0<R(\boldsymbol{\delta}\mid {\bf x})\leq 1$.

# Profile likelihood for $\sigma$
The profile likelihood for $\sigma$ is:
$$R(\sigma)=\dfrac{\max_{\mu}L(\mu,\sigma)}{L(\widehat{\mu},\widehat{\sigma})} = \left(\dfrac{\widehat{\sigma}}{\sigma}\right)^n\exp\left[\dfrac{n}{2}\left(1-\left(\dfrac{\widehat{\sigma}}{\sigma}\right)^2\right)\right].$$

The following R code shows how to obtain an approximate $95\%$ confidence interval based on this profile likelihood.

```{r}
# Delete memory
rm(list = ls())

# Simulated data from a normal with mean zero and variance 1
set.seed(123)
data = rnorm(n = 30, 0, 1 )
n = length(data)
s = sqrt(mean((data-mean(data))^2))
x.bar <- mean(data)
# MLE
MLE <- c(x.bar,s)
MLE

# Profile likelihood of sigma
R.sigma = Vectorize( function(sigma) return( (s/sigma)^n*exp( 0.5*n*(1-(s/sigma)^2)) ) )
# Visualising the profile likelihood of sigma
curve(R.sigma,0,2, n = 1000, lwd =3, col = "blue", cex.axis = 2, cex.lab = 1.5, main =  expression(paste("Profile likelihood of ", sigma), ylim = c(0,1)),
      xlab = ~sigma, ylab = "Profile")
abline(h = 0.147, lwd = 2, col = "red")
abline(v = MLE[2], lwd = 2, col = "purple", lty = 2)
# 95% Confidence interval
R.sigmaC = Vectorize( function(sigma) return(R.sigma(sigma)-0.147))

c(uniroot(R.sigmaC,c(0.6,0.8))$root,uniroot(R.sigmaC,c(1.1,1.5))$root)
```

# Profile likelihood for $\mu$
The profile likelihood for $\mu$ is:

$$R(\mu)=\dfrac{\max_{\sigma}L(\mu,\sigma)}{L(\widehat{\mu},\widehat{\sigma})} = \left(\dfrac{\widehat{\sigma}}{\widehat{\sigma}(\mu)}\right)^n = \left( \dfrac{\sum_{i=1}^n(x_i-\bar x)^2}{\sum_{i=1}^n(x_i-\mu)^2} \right)^{\frac{n}{2}}.$$
```{r}
# Delete memory
rm(list = ls())

# Simulated data from a normal with mean zero and variance 1
set.seed(123)
data = rnorm(n = 30, 0, 1 )
n = length(data)
s = sqrt(mean((data-mean(data))^2))
x.bar <- mean(data)
# MLE
MLE <- c(x.bar,s)
MLE

# Profile likelihood of mu
R.mu = Vectorize( function(mu) return(  ( sum((data-x.bar)^2)/sum((data-mu)^2) )^(0.5*n)  ))
# Visualising the profile likelihood of mu
curve(R.mu,-1,1, n = 1000, lwd =3, col = "blue", cex.axis = 2, cex.lab = 1.5, 
      main =  expression(paste("Profile likelihood of ", mu)), xlab = ~mu, ylab = "Profile", ylim = c(0,1))
abline(h = 0.147, lwd = 2, col = "red", lty = 2)
abline(v = MLE[1], lwd = 2, col = "purple", lty = 2)

# 95% Confidence interval
R.muC = Vectorize( function(mu) return(R.mu(mu)-0.147))

c(uniroot(R.muC,c(-0.8,0))$root,uniroot(R.muC,c(0.1,0.8))$root)
```
