---
title: "A Brief analysis of the Challenger data"
author: '[F. Javier Rubio](https://sites.google.com/site/fjavierrubio67/)'
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analysis of the Challenger disaster data using logistic regression

The description of this data can be found at:

[Challenger Explosion](https://www.history.com/topics/1980s/challenger-disaster)

A brief copy-paste description is as follows:

> The NASA space shuttle Challenger exploded on January 28, 1986, just 73 seconds after liftoff, bringing a devastating end to the spacecraft's 10th mission. The disaster claimed the lives of all seven astronauts aboard, including Christa McAuliffe, a teacher from New Hampshire who would have been the first civilian in space. It was later determined that two rubber O-rings, which had been designed to separate the sections of the rocket booster, had failed due to cold temperatures on the morning of the launch. The tragedy and its aftermath received extensive media coverage and prompted NASA to temporarily suspend all shuttle missions.

```{r}
# Delete memory
rm(list=ls())

# Required packages
library(knitr)

# Challenger data set
distress_ct = c(0,1,0,0,0,0,0,0,1,1,1,0,0,2,0,0,0,0,0,0,2,0,1)
temperature = c(66,70,69,68,67,72,73,70,57,63,70,78,67,53,67,75,70,81,76,79,75,76,58)
field_check_pressure = c(50,50,50,50,50,50,100,100,200,200,200,200,200,200,200,200,200,200,200,200,200,200,200)
n.rings <- rep(6,length(distress_ct))

launch <- data.frame(cbind(n.rings,distress_ct, temperature, field_check_pressure))

#launch <- launch[9:23,]

# Replacing values higher than 1 by 1 (as we are only interested in whether or not a failure happened)
#launch$distress_ct = ifelse(launch$distress_ct<1,0,1)

# Displaying the data
kable(launch)

# Fitting a logistic regression model
model <- glm(cbind(distress_ct,n.rings-distress_ct) ~temperature + field_check_pressure, family = binomial(link='logit'), data = launch)
summary(model)

#------------------------------------------------------------------
# Confidence intervals based on the profile likelihood
#------------------------------------------------------------------
kable(confint(model, level=0.95, type="LR"),digits = 4)

#------------------------------------------------------------------
# Wald Confidence intervals
#------------------------------------------------------------------
kable(confint.default(model, level=0.95),digits = 4)

#------------------------------------------------------------------
# Normal confidence intervals
#------------------------------------------------------------------
NCI <- cbind(coefficients(model)- 1.96*summary(model)$coefficients[,2],
             coefficients(model)+ 1.96*summary(model)$coefficients[,2] )

colnames(NCI) <- c("2.5 %", "97.5 %")
rownames(NCI) <- c("(Intercept)", "temperature", "pressure")
kable(NCI,digits = 4)


##################################################################################
# Brief interpretation of the model
##################################################################################

# Predicted probability of failure for the Challenger
plogis(coefficients(model)%*%c(1,31,200))

# Predicted probability of failure for minimum temperature experienced previously
plogis(coefficients(model)%*%c(1,53,200))


# The following plot presents the estimated probability 
# as a function of the temperature at pressure = 200
curve(predict(model, data.frame(temperature = x, field_check_pressure = 200),
              type = "response"),  from = 30, to = 85, ylab = "Probability of Failure", xlab = "Temperature (F)",
      cex.axis = 1.5, cex.lab = 1.5, lwd = 2, col = 'red')
```
