---
title: "Method of moments for the Kumaraswamy distribution"
author: '[Francisco Javier Rubio](https://sites.google.com/site/fjavierrubio67/)'
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Method of moments for the Kumaraswamy distribution

Let ${\bf X}=(X_1,\dots, X_n)$ be *i.i.d.* random variables with Kumaraswamy distribution with parameters $a>0$ and $b>0$. The probability density function of the Kumaraswamy distribution is
$$f(x;a,b)=abx^{{a-1}}{(1-x^{a})}^{{b-1}},\,\,\,\,\,\,{\mbox{where}}\,\,\, x\in [0,1].$$
The mean and variance of this distribution are
$$ \mu_1 = E[X] = {\frac  {b\Gamma (1+{\tfrac  {1}{a}})\Gamma (b)}{\Gamma (1+{\tfrac  {1}{a}}+b)}},$$
$$Var[X] = E[X^2] - E[X]^2 = \mu_2 -\mu_1^2 = {\frac  {b\Gamma (1+2/a)\Gamma (b)}{\Gamma (1+b+2/a)}} - \mu_1^2.$$
The MME of $a$ and $b$, $\tilde{a}$ and $\tilde{b}$, are the solution to the system of non-linear equations,
$$ \bar{{\bf X}} = {\frac  {b\Gamma (1+{\tfrac  {1}{a}})\Gamma (b)}{\Gamma (1+{\tfrac  {1}{a}}+b)}},$$
$$\frac{1}{n}\sum_{i=1}^{n}X_i^2 - \bar{{\bf X}}^2 = {\frac  {b\Gamma (1+2/a)\Gamma (b)}{\Gamma (1+b+2/a)}} - \mu_1^2.$$
which cannot be found in closed form. Thus, we can only obtain the MME using numerical methods. The following R code shows how to calculate the MME numerically.

```{r}
rm(list = ls()) # Delete memory

#------------------------------------------------------------------
# Moments of order n of the Kumaraswamy(a,b) distribution
# https://en.wikipedia.org/wiki/Kumaraswamy_distribution
#------------------------------------------------------------------
mn <- function(a,b,n){
  log.num <- log(b) + lgamma(1 + n/a) + lgamma(b)
  log.den <- lgamma(1 + b + n/a)
  return(exp(log.num-log.den))
}

#--------------------------------
# Method of Moments Estimates
#--------------------------------
library(nleqslv) # Required packages

# Function containing the implementation of the MME
# It solves a system of nonlinear equations 

MME <- function(data){
  x.bar <- mean(data)
  x2.bar <- mean(data^2)
  
  fn <- function(par){
    mom1 <- mn(exp(par[1]),exp(par[2]),1) - x.bar
    mom2 <- mn(exp(par[1]),exp(par[2]),2) - mn(exp(par[1]),exp(par[2]),1)^2 - x2.bar + x.bar^2
    return(c(mom1,mom2))
  }
  
  est <- as.vector(exp(nleqslv(c(0,0), fn)$x))
  
  return(est)
  
}

###############################################################################
# Examples
###############################################################################

#--------------------------------
# Example 1
#--------------------------------
a0 <- 1
b0 <- 1
ns <- 1000

# Simulation of Kumaraswamy distributed data
set.seed(123)
dat <- (rbeta(ns, shape1 = 1, shape2 = b0))^(1/a0)

# MME
MME(dat)

#--------------------------------
# Example 2
#--------------------------------
a0 <- 0.5
b0 <- 2
ns <- 1000

# Simulation of Kumaraswamy distributed data
set.seed(123)
dat <- (rbeta(ns, shape1 = 1, shape2 = b0))^(1/a0)

# MME
MME(dat)

#--------------------------------
# Example 3
#--------------------------------
a0 <- 5
b0 <- 3
ns <- 1000

# Simulation of Kumaraswamy distributed data
set.seed(123)
dat <- (rbeta(ns, shape1 = 1, shape2 = b0))^(1/a0)

# MME
MME(dat)
```


