---
title: 'Paired differences: the Darwin data'
author: '[Francisco Javier Rubio](https://sites.google.com/site/fjavierrubio67/)'
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# $t$-intervals: paired observations
Example taken from [1].

"Consider the data $d = (49, -67, 8, 16, 6, 23, 28, 41, 14, 29, 56, 24, 75, 60, -48)$. These were obtained by Darwin to investigate the difference between cross and self fertilization in plants. They are the differences in height between cross- and self-fertilized pairs of plants in eighths of an inch. They were used by Fisher (1991b, Chapter 3) to illustrate the use of the Student t test on the hypothesis of no difference between cross- and self-fertilized plants $H: \delta = 0$."

**References**

   1. [Statistical Inference in Science](https://www.springer.com/gb/book/9780387950198)


```{r}
# Differences between pairs
d <- c(49, -67, 8, 16, 6, 23, 28, 41, 14, 29, 56, 24, 75, 60, -48)

n <- length(d)

# t-interval
Int <- c(mean(d) - qt(0.975, df = n-1)*sqrt(var(d))/sqrt(n) , mean(d) + qt(0.975, df = n-1)*sqrt(var(d))/sqrt(n) )

library(knitr)
kable(Int, digits = 4)

# same result using the command t.test
t.test(d, conf.level = 0.95)$conf.int

```

## Controversy about misclassification in this data set

The left point of this confidence interval is very close to $0$ due to the presence of the two anomalous negative observations $-67, -48$. There is controversy about whether or not these observations were actually misclassified and the role of $X_i$ and $Y_i$ were inadvertently interchanged. Let's explore this idea by changing the sign these observations.

```{r}
# Differences between pairs
d <- c(49, 67, 8, 16, 6, 23, 28, 41, 14, 29, 56, 24, 75, 60, 48)

n <- length(d)

# t-interval
Int <- c(mean(d) - qt(0.975, df = n-1)*sqrt(var(d))/sqrt(n) , mean(d) + qt(0.975, df = n-1)*sqrt(var(d))/sqrt(n) )

library(knitr)
kable(Int, digits = 4)

# same result using the command t.test
t.test(d, conf.level = 0.95)$conf.int

```

# Test of Significance

## Original data

```{r}
# Differences between pairs
d <- c(49, -67, 8, 16, 6, 23, 28, 41, 14, 29, 56, 24, 75, 60, -48)
n <- length(d)

# Discrepancy measure (t test statistic)
t0 <- mean(d)/sqrt(var(d)/n)

# P value

P <- pt(-t0, df = n-1) + 1 - pt(t0, df = n-1)
library(knitr)
kable(P, digits = 4)

# Same result using the command t.test, fixing the alternative accordingly

t.test(d, alternative = "two.sided" )
```

## Changing the sign of two observations data

```{r}
# Differences between pairs
d <- c(49, 67, 8, 16, 6, 23, 28, 41, 14, 29, 56, 24, 75, 60, 48)
n <- length(d)

# Discrepancy measure (t test statistic)
t0 <- mean(d)/sqrt(var(d)/n)

# P value

P <- pt(-t0, df = n-1) + 1 - pt(t0, df = n-1)
library(knitr)
kable(P, digits = 8)

# Same result using the command t.test, fixing the alternative accordingly

t.test(d, alternative = "two.sided" )
```
