---
title: "Simulation of correlated covariates: implementation and a warning"
author: '[F. Javier Rubio](https://sites.google.com/site/fjavierrubio67/)'
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
bibliography: references.bib
---

# Simulating correlated covariates

When conducting simulation studies in the context of regression models (see [this link](https://rpubs.com/FJRubio/GHSim) for an example), it is common to simulate normal ``standardised covariates''. This consists of simulating normally distributed vectors, where each entry has mean zero, variance 1, and correlation coefficients $\rho_{ij}$. The correlation matrix is a symmetric (typically) positive-definite $d\times d$ matrix, with $1$s on the diagonal entries, and values $-1 < \rho_{ij} < 1$ on the off-diagonal entries. 

Simulating this sort of random vectors is simple using the R package `mvtnorm`. This only requires defining the correlation matrix as follows.

```{r}
rm(list=ls())

# Required packages
library(mvtnorm)
library(knitr)

# dimension 
d <- 3

# number of simulations
n <- 1000

# Correlation matrix
RHO <- rbind(c(1,0.5,0.75), c(0.5,1,0.25), c(0.75,0.25,1))
print(RHO)

# Simulation (each row is a simulated vector)
set.seed(123) # seed
X <- rmvnorm(n = n, mean = rep(0, d), sigma = RHO)

# sample mean and sample correlation matrix
colMeans(X)

cor(X)
```

# A warning
For simplicity, it is common to assume that the correlation of the random variables is $\rho_{ij} = \rho$ for all $i,j=1,\dots, d$. This is, the correlation matrix is:

$$
R = \left(\begin{array}{cc} 
1 & \rho & \dots & \rho\\
\rho & 1 & \dots & \rho\\
\dotso & \dots & \dots & \dots\\
\rho & \rho & \dots & 1\\
\end{array}\right)
$$
A natural question is: when is this matrix positive definite?

**A solution:**

The characteristic polynomial of $R$ is [@hoffman]

$$
P(\lambda) = \mbox{det} \left( R - \lambda I_d \right).
$$
Clearly, a root of this polynomial is $\lambda = 1-\rho$, as this value leads to a matrix $R - \lambda I_d$ with all entries equal to $\rho$, with determinant equal to zero. Then, $\lambda_1 = 1-\rho$ is an eigenvalue (and it is positive).

Moreover, the [geometric multiplicity](https://en.wikipedia.org/wiki/Eigenvalues_and_eigenvectors#Eigenspaces,_geometric_multiplicity,_and_the_eigenbasis_for_matrices) of $\lambda_1$ is : 

$$ \gamma _{R}(\lambda_1 )=n-\operatorname {rank} (R - \lambda_1 I_d).$$
Since the $R - \lambda I_d$ is a matrix with all entries equal to $\rho$, then the $\operatorname {rank} (R - \lambda_1 I_d) = 1$. Consequently, the multiplicity of $\lambda_1$ is $d-1$, and there is only one more eigenvalue. We can find the second eigenvalue by using the [property](https://en.wikipedia.org/wiki/Eigenvalues_and_eigenvectors#Additional_properties_of_eigenvalues):

$$
\begin{aligned}
d &= \operatorname {tr} (R)=\sum _{i=1}^{d}R_{ii}=\sum _{i=1}^{d}\lambda _{i}= (d-1)\lambda _{1}+\lambda _{2} \\
&= (d-1)(1-\rho) + \lambda_2
\end{aligned}
$$
Therefore, $\lambda_2 = d - (d-1)(1-\rho)$. In order for $\lambda_2$ to be positive, we need $d - (d-1)(1-\rho) >0$, which is equivalent to the condition:

$$
\rho > -\dfrac{1}{d-1}.
$$
So, when you simulate (*e.g.* normal) random variables with correlation $R$, you need to take this condition into account. Note that this condition applies to non-normal variables as well.

For $d = 2$, this implies $\rho > -1$; but for $d = 3$, you need $\rho > - 0.5$; and for $d\to \infty$, $\rho >0$.

```{r}

# dimension
d <- 10
# Correlation matrix
rho0 <- -0.25
RHO <- diag(d) + rho0*(matrix(1, nrow = d, ncol = d) - diag(d))
print(RHO)

# Determinant is negative!
det(RHO)

# So, if you try to use this correlation matrix on rmvnorm:
set.seed(123) # seed
X <- rmvnorm(n = n, mean = rep(0, d), sigma = RHO)


# dimension
d <- 10
# Correlation matrix
rho0 <- -0.1
RHO <- diag(d) + rho0*(matrix(1, nrow = d, ncol = d) - diag(d))
print(RHO)

# Determinant is positive
det(RHO)

```


# References
