---
title: 'Logistic Regression: The Beetle data set'
author: '[F. Javier Rubio](https://sites.google.com/site/fjavierrubio67/)'
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Maximum likelihood estimation in the logistic regression model

This real data example illustrates the method of maximum likelihood estimation for the parameters of the logistic regression model. The R code compares the results using the following methods:

* The Newton's method.
* A direct implementation. Implementing the log-likelihood function and maximising it using the command `optim()`.
* Using the command `glm()`.

The results are virtually the same, as expected.

```{r}
# Delete Memory
rm(list=ls())

# Required packages
library(knitr)
library(plotly)
## Warning: package 'knitr' was built under R version 3.4.3
# Beetle data
dat =  cbind(c(1.6907, 1.7242, 1.7552, 1.7842, 1.8113,1.8369, 1.861, 1.8839), 
             c(59, 60, 62, 56, 63, 59, 62, 60), 
             c(6, 13, 18, 28, 52, 53, 61, 60))

# Naming the columns
colnames(dat) <- c("Dose", "N","x")

# Displaying the data
kable(dat, digits = 4)

ns = length(dat[,1]) # number of responses
ni <- dat[,2] # numbers of beetles
yi <- dat[,3] # numbers of dead beetles after exposure
xi <- dat[,1] # Dose

###############################################################################################################
# Maximum Likelihood Estimation using three methods
###############################################################################################################

#--------------------------
# Using  Newton's method
#--------------------------
theta = c(0,0) # Initial value

N = 100 # Number of iterations 

iter <- matrix(0,nrow=N,ncol=2)

# Running the iterations
for(i in 1:N){
  iter[i,] <- theta
  pi <- exp(theta[1] + theta[2]*xi)/( 1 + exp(theta[1] + theta[2]*xi) )
  S <- c( sum(yi - ni*pi), sum(xi*(yi - ni*pi)))
  I <- rbind( c(sum(ni*pi*(1-pi)), sum(ni*xi*pi*(1-pi)) ), c(sum(ni*xi*pi*(1-pi)), sum(ni*xi^2*pi*(1-pi)) ) )
  theta <- theta + solve(I)%*%S
  
}

# Illustrating the iterations
plot(iter[1:10,], type = "p")
head(iter)

df <- data.frame(x = iter[,1], y = iter[,2], f = iter[,2])
p <- df %>%
  plot_ly(
    x = ~x,
    y = ~y,
    frame = ~f,
    type = 'scatter',
    mode = 'markers',
    showlegend = F,
    size = 40
  )

p

# MLE after 100 iterations
theta

# LD50
-theta[1]/theta[2]

#--------------------------
# Direct implementation
#--------------------------

# log-likelihood function
lpl = function(par){
  var <- vector()
  for(i in 1:ns) var[i] = dat[i,3]*log(plogis(par[1]+par[2]*dat[i,1])) + (dat[i,2]-dat[i,3])*log(1-plogis(par[1]+par[2]*dat[i,1]))
  return(-sum(var))
}

# Optimisation step using optim
OPT <- optim(c(-60,30),lpl,control=list(maxit=20000,abstol=1e-18,reltol=1e-18))
# MLE
OPT$par

#--------------------------
# Using glm
#--------------------------
glm(cbind(yi,ni-yi) ~ xi, family=binomial(logit), data=data.frame(dat))


###############################################################################################################
# Visualisation of the fitted dose-response model
###############################################################################################################

# Proportions
propi <- yi/ni

# Fitted logistic model
fit.logis <- Vectorize(function(d)  plogis(theta[1] + theta[2]*d))

curve(fit.logis,1.6,2, xlab = "Dose", ylab = "Proportion", main = "Fitted Model", lwd = 2, cex.axis = 1.5, cex.lab = 1.5)
points(xi,propi,pch=10,col="red",lwd=2)

```
