---
title: "$t$-intervals: unpaired observations"
author: '[Francisco Javier Rubio](https://sites.google.com/site/fjavierrubio67/)'
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Life spans of wild type vs. transgenic mosquitoes

Example based on [Joseph C. Watkins](https://www.math.arizona.edu/~jwatkins/).  The life span in days of 88 wildtype and 99 transgenic mosquitoes is given in the following data set.

## $t-$interval assuming equal variance

```{r}
# Required package
library(ggplot2)

# Reading the data online
mosquitoes<-read.delim("https://www.math.arizona.edu/~jwatkins//mosquitoes.txt")

# Displaying the data
# See also: http://rpubs.com/FJRubio/TwoSamplesVT
boxplot(mosquitoes)

# First sample
X <- mosquitoes[!is.na(mosquitoes[,1]),1]
m <- length(X)
m
var(X)
summary(X)

# Second sample
Y <- mosquitoes[!is.na(mosquitoes[,2]),2]
n <- length(Y)
n
var(Y)
summary(Y)

# Violin plots of the two populations
df <- reshape2::melt(data.frame(cbind(X,Y)), id.vars = NULL)
vp <- ggplot(df, aes(x = variable, y = value)) + geom_violin(scale="width",adjust = 1,width = 0.5,fill = "grey80") + geom_boxplot(width=0.075,fatten = 3) +
  theme_bw() + xlab("Data") + ylab("Density") + theme(axis.text.x = element_text(size = 14), axis.text.y = element_text(size = 14)) +
  theme(axis.title.x = element_text(size = 14), axis.title.y = element_text(size = 14)) + 
  stat_summary(fun.y = mean, colour = c("red","blue"), geom="point", shape=18, size=4) +
  scale_x_discrete(labels = c("Wildtype", "Transgenic"))
vp 

# Pooled variance
Sp2 <- ((m-1)*var(X) + (n-1)*var(Y))/(m+n-2)
Sp2

# t-interval : equal variance

Int <- c( mean(X) - mean(Y) - qt(0.99,df=m+n-2)*sqrt(Sp2)*sqrt(1/m+1/n), mean(X) - mean(Y) + qt(0.99,df=m+n-2)*sqrt(Sp2)*sqrt(1/m+1/n))

library(knitr)
kable(Int,digits = 4)
```

## $t-$interval using Welch and Satterthwaite's approximation
```{r}
library(knitr)
# Reading the data online
mosquitoes<-read.delim("https://www.math.arizona.edu/~jwatkins//mosquitoes.txt")

# Displaying the data
boxplot(mosquitoes)

# First sample
X <- mosquitoes[!is.na(mosquitoes[,1]),1]
m <- length(X)

# Second sample
Y <- mosquitoes[!is.na(mosquitoes[,2]),2]
n <- length(Y)

# Effective degrees of freedom
nu <- (var(X)/m + var(Y)/n)^2/( var(X)^2/(m^2*(m-1)) + var(Y)^2/(n^2*(n-1)) )
kable(nu,digits = 4)

# Studentised variance
SV <- var(X)/m + var(Y)/n

# t-interval : equal variance

Int <- c( mean(X) - mean(Y) - qt(0.975,df=nu)*sqrt(SV), mean(X) - mean(Y) + qt(0.975,df=nu)*sqrt(SV))

kable(Int,digits = 4)


# Calculating this same interval using the command t-test
t.test(X,Y,conf.level=0.95)$ conf.int
```

Both approaches give similar results in this case, although the second approach produces a confidence interval that is farther from zero.

# Test of Significance

Consider testing a difference in the mean life span. This is

$$H_0: \mu_1 =  \mu_2 \,\,\,\,\,\, vs. \,\,\,\,\,\, H_1: \mu_1 \neq \mu_2$$


```{r}
library(knitr)
# Reading the data online
mosquitoes<-read.delim("https://www.math.arizona.edu/~jwatkins//mosquitoes.txt")

# Displaying the data
boxplot(mosquitoes)

# First sample
X <- mosquitoes[!is.na(mosquitoes[,1]),1]
m <- length(X)

# Second sample
Y <- mosquitoes[!is.na(mosquitoes[,2]),2]
n <- length(Y)

# Effective degrees of freedom
nu <- (var(X)/m + var(Y)/n)^2/( var(X)^2/(m^2*(m-1)) + var(Y)^2/(n^2*(n-1)) )
kable(nu,digits = 4)

# Studentised variance
SV <- var(X)/m + var(Y)/n

# t-statistic
t0 <- (mean(X) - mean(Y))/sqrt(SV)


# P-value
P <- pt(-t0 , df = nu) + 1-pt(t0 , df = nu)
kable(P,digits = 4)

# Calculating the p-value the command t-test
t.test(X,Y,conf.level=0.95, alternative = "two.sided")

```


We now want to test if the transgenic modification of the mosquitoes reduces the mean life span. This is

$$H_0: \mu_1 \leq  \mu_2 \,\,\,\,\,\, vs. \,\,\,\,\,\, H_1: \mu_1 > \mu_2$$


```{r}
library(knitr)
# Reading the data online
mosquitoes<-read.delim("https://www.math.arizona.edu/~jwatkins//mosquitoes.txt")

# First sample
X <- mosquitoes[!is.na(mosquitoes[,1]),1]
m <- length(X)

# Second sample
Y <- mosquitoes[!is.na(mosquitoes[,2]),2]
n <- length(Y)

# Effective degrees of freedom
nu <- (var(X)/m + var(Y)/n)^2/( var(X)^2/(m^2*(m-1)) + var(Y)^2/(n^2*(n-1)) )
kable(nu,digits = 4)

# Studentised variance
SV <- var(X)/m + var(Y)/n

# t-statistic
t0 <- (mean(X) - mean(Y))/sqrt(SV)


# P-value
P <-  1-pt(t0 , df = nu)
kable(P,digits = 4)

# Calculating the p-value the command t-test
t.test(X,Y,conf.level=0.95, alternative = "greater")
```

